---
title: "Interactive Data Exploration"
author: "Jordin Shurpin"
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(plotly)
library(DT)

softball <- read_csv("softball_divisions.csv")
```

## Explore the Data: Control What You See

Welcome to the **Exploration** page! Here, you can interact directly with the collegiate softball data. 

- **In the interactive chart below:** Hover over any point to see the team's name, division, home runs, and win rate. Click divisions in the legend on the right to toggle them on or off, or double-click one to view it exclusively.
- **In the interactive table further down:** Search for your favorite school, sort by any column, and filter across divisions.

### Interactive Scatter: Home Run Rate vs. Win Percentage

```{r}
#| label: interactive-plot
#| fig-height: 6

p <- ggplot(
  softball, 
  aes(
    x = hr_pct, 
    y = win_pct, 
    color = division,
    text = paste0(
      "<b>", team, "</b> (", division, ")<br>",
      "Win %: ", scales::percent(win_pct, accuracy = 0.1), "<br>",
      "Record: ", wins, "-", losses, "<br>",
      "HR %: ", scales::percent(hr_pct, accuracy = 0.01), "<br>",
      "Total HRs: ", hr
    )
  )
) +
  geom_point(alpha = 0.7, size = 2) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.5)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_color_brewer(palette = "Set1") +
  labs(
    x = "Home Run Percentage (HR / AB)",
    y = "Season Win Percentage",
    color = "Division"
  ) +
  theme_minimal(base_size = 12)

ggplotly(p, tooltip = "text")
```

---

### Searchable & Filterable Team Table

Filter by division, search for any team, or click any column header to sort.

```{r}
#| label: interactive-table

softball |> 
  select(
    Division = division,
    Team = team,
    Games = games,
    Wins = wins,
    Losses = losses,
    `Win %` = win_pct,
    `At-Bats` = ab,
    `Home Runs` = hr,
    `HR %` = hr_pct
  ) |> 
  mutate(
    `Win %` = scales::percent(`Win %`, accuracy = 0.1),
    `HR %` = scales::percent(`HR %`, accuracy = 0.01)
  ) |> 
  datatable(
    filter = "top",
    options = list(
      pageLength = 15,
      autoWidth = TRUE,
      dom = "Bfrtip"
    ),
    rownames = FALSE
  )
```
