---
title: "Interactive Hitter Explorer"
author: "Jordin Shurpin"
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(plotly)
library(DT)

batting <- read_csv("batting2022.csv") |>
  filter(PA >= 100)
```

## Explore Division I Hitters

Use this interactive page to search, filter, and inspect **2,175 individual NCAA Division I hitters** from the 2022 season (qualified with at least 100 plate appearances).

### Interactive Power vs. On-Base Scatter

Hover over any point to view the player's name, school, home runs, and slash line (BA / OBP / SLG). You can also zoom into clusters of hitters or isolate specific home run tiers.

```{r}
#| label: interactive-hitters-plot
#| fig-height: 6

p_hitters <- ggplot(
  batting, 
  aes(
    x = OBP, 
    y = SLG, 
    color = HR,
    text = paste0(
      "<b>", Player, "</b> (", Team, ")<br>",
      "Slash: ", scales::number(BA, accuracy = 0.001), " / ",
      scales::number(OBP, accuracy = 0.001), " / ",
      scales::number(SLG, accuracy = 0.001), "<br>",
      "OPS: ", scales::number(OPS, accuracy = 0.001), "<br>",
      "HR: ", HR, " | RBI: ", RBI, "<br>",
      "PA: ", PA, " (AB: ", AB, ")"
    )
  )
) +
  geom_point(alpha = 0.7, size = 1.8) +
  scale_color_viridis_c(option = "plasma") +
  scale_x_continuous(labels = scales::number_format(accuracy = 0.01)) +
  scale_y_continuous(labels = scales::number_format(accuracy = 0.01)) +
  labs(
    x = "On-Base Percentage (OBP)",
    y = "Slugging Percentage (SLG)",
    color = "Home Runs"
  ) +
  theme_minimal(base_size = 12)

ggplotly(p_hitters, tooltip = "text")
```

### Searchable Hitter Database

Search for specific athletes or universities, sort columns by clicking headers, and page through the complete Division I dataset.

```{r}
#| label: interactive-hitters-table

batting |>
  select(Player, Team, GP, PA, AB, HR, RBI, BA, OBP, SLG, OPS) |>
  arrange(desc(HR)) |>
  datatable(
    filter = "top",
    options = list(
      pageLength = 10,
      autoWidth = TRUE,
      dom = "Bfrtip"
    ),
    rownames = FALSE,
    colnames = c("Player", "Team", "GP", "PA", "AB", "HR", "RBI", "AVG", "OBP", "SLG", "OPS")
  ) |>
  formatRound(columns = c("BA", "OBP", "SLG", "OPS"), digits = 3)
```
