---
title: "Individual Hitters in College Softball"
author: "Jordin Shurpin"
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| message: false

library(tidyverse)
```

## Individual Hitting Performance: Power vs. On-Base Ability

Welcome to our college softball analytics website! Before analyzing team records and divisional differences, we begin with the fundamental building blocks of softball offense: **the individual hitters**.

Below are **2,175 individual NCAA Division I hitters** from the 2022 season (minimum 100 plate appearances). We examine the relationship between on-base ability (OBP) and power (Slugging Percentage, SLG), shaded by total home runs hit.

```{r}
#| label: hitters-plot
#| fig-width: 9
#| fig-height: 6
#| message: false
#| warning: false

batting <- read_csv("batting2022.csv")

batting |>
  filter(PA >= 100) |>
  ggplot(aes(x = OBP, y = SLG, color = HR)) +
  geom_point(alpha = 0.65, size = 1.9) +
  geom_smooth(
    method = "lm", 
    se = FALSE, 
    color = "darkred", 
    linetype = "dashed", 
    linewidth = 0.8
  ) +
  scale_color_viridis_c(option = "plasma") +
  scale_x_continuous(labels = scales::number_format(accuracy = 0.01)) +
  scale_y_continuous(labels = scales::number_format(accuracy = 0.01)) +
  labs(
    title = "Batting Power vs. On-Base Ability in NCAA Division I Softball",
    subtitle = "2,175 individual hitters with at least 100 plate appearances (2022 season)",
    x = "On-Base Percentage (OBP)",
    y = "Slugging Percentage (SLG)",
    color = "Home Runs",
    caption = "Source: SCORE Sports Data Repository / NCAA Softball"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )
```

### Key Takeaways from Individual Data

- **Power and on-base skills rise together:** Hitters who get on base at elite rates (.400+ OBP) also tend to slug significantly higher (.600+ SLG).
- **The impact of the long ball:** Players in the highest home run tier (bright yellow points, 15+ HRs) dominate the upper-right quadrant of offensive production.
- **From individual hitters to team success:** While individual hitters drive run production, does stacking home run power actually guarantee team victories across different divisions of college softball?

Explore our full findings across the site:

- Check out the **[Story](story.qmd)** tab for a division-by-division analysis (D1, D2, D3, and JUCO) comparing home run rates to game outcomes.
- Use the **[Exploration](exploration.qmd)** tab to interactively filter teams and examine the data yourself.
